---
title: "Student's t-Distribution"
author: "RQuesT"
date: "`r Sys.Date()`"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
```

## 1. Executive Overview

* **The Core Purpose**:
  * Designed for population mean inferences when true population variance ($\sigma^2$) is unknown.
  * Replaces standard normal $Z$-distribution when estimating spread from small samples.

* **Historical Origins**:
  * Formulated by chemist **William Sealy Gosset** in 1908.
  * Published under the famous pseudonym **"Student"** while at Guinness Brewery.

---

## 2. Fundamental Mathematical Properties 

* **Definition & Statistic**:
  * Calculated as $t = \frac{\bar{x} - \mu}{s / \sqrt{n}}$ for samples from normal populations.
  * Uses sample standard deviation ($s$) to estimate unknown population standard deviation ($\sigma$).

* **Key Geometric Traits**:
  * **Symmetrical**: Perfectly balanced around a mean of zero.
  * **Continuous**: Spans infinitely from $-\infty$ to $+\infty$.
  * **Heavier Tails**: Less peaked at the center with thicker tails than the $Z$-distribution.
  * **Wider Critical Values**: Thicker tails adjust for small-sample sampling uncertainty.

* **Variance & Degrees of Freedom**:
  * Governed by **degrees of freedom** ($df = n - 1$).
  * Variance exceeds 1: Calculated as $\frac{df}{df - 2}$ for $df > 2$ (or $\frac{n - 1}{n - 3}$ for $n > 3$).
  * **Convergence**: Approaches the standard normal $Z$-distribution as $df \to \infty$.

---

## 3. Practical & Clinical Perspectives 

* **Small-Sample Adjustment**:
  * $Z$-distribution approximations degrade as sample sizes shrink.
  * Inflates critical thresholds to prevent false positive conclusions.

* **Degrees of Freedom Mechanics**:
  * Calculated as $df = n - 1$.
  * 1 degree of freedom lost because sample mean fixes central position.
  * Analogy: Like fence post gaps—fixing the mean leaves $n - 1$ independent values.

* **Operational Rules & Reporting**:
  * **Preferred**: Highly recommended for sample sizes $n < 60$.
  * **Essential**: Critical for small samples with $n \le 30$.
  * **Avoid Dichotomania**: Rejecting rigid $P \le 0.05$ pass/fail cutoffs.
  * **Complete Reporting**: Always report $t$-statistic, $df$, $P$-value, and 95% Confidence Intervals together.

---

## 4. Visualizing the $t$-Distribution vs. $Z$-Distribution

```{r t-vs-z-plot, fig.width=8, fig.height=4.5, fig.align='center'}
# Visualizing Convergence and Tail Thickness
x_vals <- seq(-4, 4, length.out = 400)
df_data <- data.frame(
  x = rep(x_vals, 3),
  Density = c(dnorm(x_vals), dt(x_vals, df = 3), dt(x_vals, df = 10)),
  Distribution = factor(rep(c("Standard Normal (Z)", "t (df = 3)", "t (df = 10)"), 
                            each = 400),
                        levels = c("Standard Normal (Z)", "t (df = 10)", "t (df = 3)"))
)

ggplot(df_data, aes(x = x, y = Density, color = Distribution, linetype = Distribution)) +
  geom_line(size = 1.1) +
  scale_color_manual(values = c("black", "#2B5C8F", "#D95F02")) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted")) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Student's t-Distribution vs. Standard Normal (Z)",
    subtitle = "Thicker tails at lower df converge to Z as df increases",
    x = "Statistic (t or Z)",
    y = "Probability Density"
  ) +
  theme(legend.position = "top")
```

---

## 5. Summary

* **t distribution is**:
  * A mathematical tool for analyzing small groups of data.
  * Solves the problem of unknown population spread.

* **How it works**:
  * Has a flatter top and wider sides (thicker tails) than a normal bell curve.
  * Wider tails inflate critical boundaries, preventing false conclusions from small samples.

* **The Magic of Sample Size**:
  * Curve shape is controlled by **degrees of freedom** ($df = n - 1$).
  * As sample size grows, extra uncertainty fades away.
  * The $t$-distribution transforms until it perfectly matches the standard normal bell curve.

---

## 6. R Computation

```
# 1. Calculating Critical t-Value for 95% Confidence (df = 29)
qt(p = 0.975, df = 29)
[1] 2.04523
# 2. Calculating Two-Tailed P-Value for t = 2.15 (df = 18)
2 * (1 - pt(q = 2.15, df = 18))
[1] 0.04539693
# 3. Running a One-Sample t-Test in R
# t.test(x = sample_data, mu = hypothesized_mean)
```

---

## References

* Daniel, W. W., & Cross, C. L. *Biostatistics: A Foundation for Analysis in the Health Sciences* (10th ed.) .
* Campbell, M. J. *Statistics at Square One* (12th ed.).
