---
title: "Descriptive Statistics"
author: "Kaung Myat Khant"
date: "`r Sys.Date()`"
output: 
    powerpoint_presentation:
        keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = F,
                      warning = F,
                      comment = "")
```

## Outline  

1.  Classify data and check errors or missing values  

2.  Visualize and explore the data  

3.  Measure Centrl tendency  

4.  Measure Dispersion  

5.  Outliers  

## Classify data  

-   Numerical or Categorical  

-   Numerical --> Continuous or Discrete    

-   Categorical --> Nominal or Ordinal    

## Error checking  

-   Categorical data --> Category check

-   Numerical data --> Range check  

## Missing values  

-   Check the missing codes [NAs or user defined NA or tagged NA like -9,99]  

-   Investigate the reason for missing  
    -   Completely at random  
    -   At random  
    -   Not at random  
    
## Data visualization and exploring data  

-   Numerical variables  
    -   Histogrm   
    -   Boxplot

-   Categorical variables  
    -   Bar graph  
    -   Pie chart  
    
##  Measures of centrl tendency  

-   Numerical variable 
    -   if symmetric without extreme value --> arithmetic mean  
    -   if positive skewed --> geometric mean  
    -   if outliers pull the mean --> trimmed mean  
    -   Median is more robust to skewed data

##  Measures of centrl tendency   

-   Ordinal variable  
    -   mode or median  
    -   mean for comparing two group  
    
-   Nominal variable --> mode  

-   Arithmetic mean is useful for inferential statistics or total 

## Measures of dispersion  

-   Range shows the total spread: minimum and maximum  

-   Standard deviation shows average spread for the normally distributed data  
    -   2SD - mean + 2SD is 95% of data for normal distribution

-   Interquartile range show spread of central 50%  of data
    -   Inter-decile range shows middle 80% of data

-   Coefficient of variation CV shows relative variability  

## Outlier  

-   Can be seen in box-plot and stem and leaf plot  

-   Beyond 1.5 times IQR from the Q1 & Q3  

-   Needs investigation  
    -   natural outlier or errors  
    
    
```{r data, include=FALSE}
sbp <- c(110, 112, 115, 120, 122, 122, 124, 125, 128, 130, 130, 132, 135, 138, 140, 142, 145, 150, 155, 175)
library(tidyverse)
library(patchwork)
```

## Systolic Blood Pressure (SBP) in 20 Patients (mmHg) after a 4-week trial of Medication X  

```{r Dataframe, include=TRUE}
df <- data.frame(sbp =sbp)
df |> slice_head(n =10) |> 
    knitr::kable()
```

## Histograms

```{r hist, include=TRUE}
binsturge <- ggplot(df)+geom_histogram(aes(sbp), binwidth = 10.83)
bin_num1 <- ggplot(df)+geom_histogram(aes(sbp), bins = 30)
bin_num2 <- ggplot(df)+geom_histogram(aes(sbp), bins = 3)
binsturge+bin_num1+bin_num2
```

## Boxplot
```{r box, include=TRUE}
ggplot(df)+geom_boxplot(aes(sbp))
```

## Stem and leaf
```{r stem, include=TRUE}
stem(df$sbp) 
```

## Summary table
```{r tab, include=TRUE}
df |> 
    summarise(
        Mean = mean(sbp),
        Median = median(sbp),
        SD = sd(sbp),
        IQR = IQR(sbp),
        Q1 = quantile(sbp, 0.25),
        Q3 = quantile(sbp,0.75)
    ) |> 
    pivot_longer(cols = everything(),
                 names_to = "Measures",
                 values_to = "SBP") |> 
    knitr::kable(digits = 2)
```

