---
title: "The F-Distribution & One-Way ANOVA"
subtitle: "A Concise Guide to Variance Ratios, ANOVA, and Post-Hoc Comparisons"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
```

## 1. Executive Summary & Core Concept

* **Statistical Foundation**: Developed by Sir Ronald A. Fisher in the 1920s.
* **Core Function**: Measures the ratio of two independent sample variances.
* **Primary Purpose**: Evaluates variability across multiple groups simultaneously.
* **Key Metric**: The **Variance Ratio** ($F = \text{Between-Group Variance} / \text{Within-Group Variance}$).
* **Main Advantage**: Controls family-wise Type I error rates across $k$ groups.

---

## 2. Mathematical Definition & Key Properties

* **Ratio of Chi-Squares**: Ratio of two independent chi-square variables divided by their degrees of freedom.
* **Parameter Pair**: Governed by numerator $df_1 = n_1 - 1$ and denominator $df_2 = n_2 - 1$.
* **Non-Negative Support**: Always $F \ge 0$ since variances are non-negative.
* **Asymmetric Shape**: Positively skewed with a long right tail.
* **Reciprocal Property**: Lower percentiles satisfy $F_{\alpha, df_1, df_2} = 1 / F_{1-\alpha, df_2, df_1}$.
* **Special Equivalence**: For $k = 2$ groups, the $F$-statistic equals Student's $t^2$ ($F = t^2$).

---

## 3. Behavior & Interpretation

* **Null Scenario ($F \approx 1$)**:
  * Between-group variation matches background noise.
  * Group mean differences are likely due to random chance.
* **Alternative Scenario ($F \gg 1$)**:
  * Between-group variation significantly exceeds random variation.
  * Signals genuine treatment or group differences.
* **Underlying Assumptions**:
  * Normally distributed populations.
  * Homogeneity of variance (equal group variances).
  * Independent sampling of observations.

---

## 4. Scope of Applications

* **Variance Ratio Test**: Compares two population variances ($H_0: \sigma_1^2 = \sigma_2^2$).
* **One-Way ANOVA**: Tests equality of $k$ group means ($H_0: \mu_1 = \mu_2 = \dots = \mu_k$).
* **Advanced Designs**: Applies to Randomized Complete Block, Factorial, and Repeated Measures ANOVA.
* **Multivariable Regression**: Tests overall model significance ($F = MSR / MSE$) and extra sums of squares.

---

## 5. One-Way ANOVA Hypothesis Testing Example

### Study Design & Raw Data
* **Context**: Testing 3 weight-loss diets ($k = 3$) across $N = 9$ subjects ($n_j = 3$ per diet) .
* **Diet 1 ($Group_1$)**: 4, 5, 6 pounds lost.
* **Diet 2 ($Group_2$)**: 7, 8, 9 pounds lost.
* **Diet 3 ($Group_3$)**: 10, 11, 12 pounds lost.

### Step-by-Step Calculation Summary
* **Group Means**: $\bar{x}_1 = 5$, $\bar{x}_2 = 8$, $\bar{x}_3 = 11$.
* **Grand Mean**: $\bar{x}_{..} = (5 + 8 + 11) / 3 = 8$.
* **Sum of Squares Among ($SSA$)**: $3(5-8)^2 + 3(8-8)^2 + 3(11-8)^2 = 54$.
* **Among Degrees of Freedom ($df_A$)**: $k - 1 = 3 - 1 = 2$.
* **Mean Square Among ($MSA$)**: $SSA / df_A = 54 / 2 = 27$.
* **Sum of Squares Within ($SSW$)**: $(1+0+1) + (1+0+1) + (1+0+1) = 6$.
* **Within Degrees of Freedom ($df_W$)**: $N - k = 9 - 3 = 6$.
* **Mean Square Within ($MSW$)**: $SSW / df_W = 6 / 6 = 1$.
* **Calculated $F$-Statistic**: $F = MSA / MSW = 27 / 1 = 27$.

### ANOVA Summary Table

| Source of Variation | Sum of Squares ($SS$) | Degrees of Freedom ($df$) | Mean Square ($MS$) | $F$-Value |
| :--- | :---: | :---: | :---: | :---: |
| **Among Groups** | 54 | 2 | $MSA = 27$ | **$F = 27.0$** |
| **Within Groups** | 6 | 6 | $MSW = 1$ | — |
| **Total** | 60 | 8 | — | — |

* **Decision**: Computed $F = 27$ far exceeds the critical value $F_{0.05, 2, 6} = 5.14$ ($P < 0.001$), confirming statistically significant differences among the three diets.

---

## 6. Post-Hoc Pairwise Comparisons (Tukey HSD & Bonferroni)

* **Why Post-Hoc Testing Is Essential**:
  * An overall $F$-test indicates *that* group means differ, but not *which* specific pairs differ.
  * Running multiple unadjusted $t$-tests inflates the overall Type I error rate.
  * Post-hoc tests pinpoint specific pairwise differences while maintaining error control.

* **Tukey's Honestly Significant Difference (HSD)**:
  * **Method**: Uses the Studentized Range distribution to calculate a single critical threshold for all pairwise differences.
  * **Best Used For**: All possible pairwise comparisons among group means.
  * **Key Feature**: Maintains exact family-wise error rate $\alpha$ without being overly conservative.
  * **In R**: Evaluated via `TukeyHSD(anova_model)`.

* **Bonferroni Adjustment**:
  * **Method**: Adjusts significance threshold to $\alpha / m$ (or multiplies $p$-values by $m$, where $m = k(k-1)/2$ comparisons).
  * **Best Used For**: Small set of pre-planned or custom pairwise comparisons.
  * **Key Feature**: Highly robust and versatile, but can become overly conservative as $k$ grows.
  * **In R**: Evaluated via `pairwise.t.test(..., p.adjust.method = "bonferroni")`.

---

## 7. Complete R Code Implementation

```{r implementR1, include=TRUE}
# --- 1. Load Study Data ---
diet_data <- data.frame(
  Diet = factor(rep(c("Diet 1", "Diet 2", "Diet 3"), each = 3)),
  WeightLoss = c(4, 5, 6, 7, 8, 9, 10, 11, 12)
)

# --- 2. Fit One-Way ANOVA Model ---
anova_model <- aov(WeightLoss ~ Diet, data = diet_data)

# --- 3. Display ANOVA Summary Table ---
summary(anova_model)
```

```{r fDistribution, include=TRUE}
# --- 4. Plot F-Distribution Curve & Test Statistic ---
f_stat <- summary(anova_model)[[1]][["F value"]][1]
df1 <- summary(anova_model)[[1]][["Df"]][1]
df2 <- summary(anova_model)[[1]][["Df"]][2]

x_vals <- seq(0, 30, length.out = 500)
y_vals <- df(x_vals, df1, df2)
plot_df <- data.frame(x = x_vals, y = y_vals)
criticalF <- qf(p = 1-0.05, df1 = 2, df2 = 6)
ggplot(plot_df, aes(x = x, y = y)) +
  geom_line(color = "navy", size = 1) +
  geom_vline(xintercept = f_stat, color = "forestgreen", linetype = "solid", size = 1) +
  geom_vline(xintercept = criticalF, color = "firebrick", linetype = "dashed", size = 1) +
  labs(
    title = "F-Distribution Curve (df1 = 2, df2 = 6)",
    subtitle = paste("Calculated F-Statistic(green) =", f_stat, "(P < 0.001),",
                     "Critical F-value(firecolor) =", round(criticalF,2), "(alpha =0.05)"),
    x = "F Value",
    y = "Probability Density"
  ) +
  theme_minimal()
```


```{r postHoc, include=TRUE}
# --- 5. Post-Hoc Comparisons: Tukey HSD ---
cat("\n--- Tukey HSD Pairwise Comparisons ---\n")
tukey_results <- TukeyHSD(anova_model)
print(tukey_results)

# --- 6. Post-Hoc Comparisons: Bonferroni Adjustment ---
cat("\n--- Bonferroni Adjusted Pairwise t-Tests ---\n")
bonf_results <- pairwise.t.test(
  x = diet_data$WeightLoss,
  g = diet_data$Diet,
  p.adjust.method = "bonferroni"
)
print(bonf_results)
```
