---
title: "Probability Distributions"
author: "Kaung Myat Khant"
date: "`r Sys.Date()`"
output: 
    powerpoint_presentation:
        keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = FALSE)
```

## What is a probability distribution?

-   Mathematical model that characterizes a *population*

-   Describes all possible values of a random variable

-   Describes likelihood associated with each value

## Main categories of probability distribution

1.  Discrete probability distributions
    -   Binomial distribution

    -   Poisson distribution
2.  Continuous probability distributions
    -   Normal distribution

    -   t distribution

    -   F distribution

    -   Chi square distribution

## Two rules of a valid probability distribution

-   Probability of any event cannot be negative

-   Total sum of all probabilities = 1 (Total area under the curve = 1)

## Binomial distribution

-   Describes the number of successes in a fixed number of trial (n)

-   Each trial has only two *mutually exclusive* outcomes (p,1-p)

-   Used in analysing proportions or risk (dead/alive; treatment success/fail)

-   *Probability* of success (p) must be *constant* for every trial

-   *Outcome* of one trial must be *independent* of the result of another trial

## Parameters of binomial distribution

-   Mean (measure of central tendency) is $\mu = n \times p$

-   Variance (measure of spread) is $\sigma^2 = n \times p \times (1-p)$

-   Parameters are *number of trials (n)* and *probability of success (p)*

## Binomial distribution formuala

$$P(X = x) = {}_nC_x \cdot p^x \cdot (1-p)^{n-x}$$

## Example of binomial distribution

-   Let's flip a fair coin for 10 times and the probability of head is *p*

-   p = 0.5; n = 10

-   $\mu = p \times n = 5$

-   $\sigma^2 = n \cdot p \cdot (1-p) = 2.5$ ; $\sigma = 1.581$

```{r setBinom, include=FALSE}
# Load library
library(ggplot2)

# 1. Define data in a data frame
n <- 10
p <- 0.5
df <- data.frame(
    x = 0:n,
    y = dbinom(0:n, size = n, prob = p)
)
```

## Data

```{r dataBinom, include=TRUE}
head(df, 11) |> knitr::kable()
```


## Binomial Distribution  

```{r curveBinom, include=TRUE}
ggplot(df, aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "steelblue", color = "white") +
    geom_vline(xintercept = 5, color = "red", linetype = "dashed", size = 1) +
    geom_vline(xintercept = c(5 - 1.6, 5 + 1.6), 
               color = "darkgreen", linetype = "dotted", size = 1) +
    labs(title = paste("Binomial Distribution Curve of flipping a fair coin \n (n =", n, ", p =", p, ")"),
         subtitle = paste("Mean (Red):", 5, "| Std Dev (Green):", 1.6),
         x = "Number of Heads",
         y = "Probability") +
    scale_x_continuous(labels = 0:n, breaks = 0:n)+
    theme_minimal()
```

## Binomial distribution of an unfair coin

```{r unfairCoin, include=TRUE}
n <- 10
p <- 0.8
df <- data.frame(
    x = 0:n,
    y = dbinom(0:n, size = n, prob = p)
)


ggplot(df, aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "steelblue", color = "white") +
    labs(title = paste("Binomial Distribution Curve of flipping an unfair coin  \n (n =", n, ", p =", p, ")"),
         x = "Number of Heads",
         y = "Probability") +
    scale_x_continuous(labels = 0:n, breaks = 0:n)+
    theme_minimal()
```

