---
title: "Tidy data"
author: "KMK"
date: "2023-10-24"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = "")
knitr::opts_knit$set(root.dir = "d:/docs/Burnet Statistics training/output")
```

```{r wd, include=F}
getwd()
```

## Tidy data

> ***"Happy families are all alike;every unhappy family is unhappy in its own way."*** --- **Leo Tolstoy**

> ***Tidy datasets are all alike but every messy dataset is messy in its own way.***

## Tidy data: A standardized way to link structure of a dataset with its semantics
- 80% of data analysis is spent on cleaning and preparing data


## Tidy data: A standardized way to link structure of a dataset with its semantics
- Data structure
    - made up of rows and columns
    - columns almost always labelled



```{r datastructure, include=F}
person <- c("John Smith","Jane Doe","Mary Johnson")
treatmenta <- c("_","16","3")
treatmentb <- c("2","11","1")
tab1 <- data.frame(person,treatmenta,treatmentb)
treatment <- c("treatmenta","treatmentb")
`John Smith` <- c("_","2")
`Jane Doe` <- c("16","11")
`Mary Johnson` <- c("3","1")
tab2 <- data.frame(treatment,`John Smith`,`Jane Doe`,`Mary Johnson`)
name <- rep(c("John Smith","Jane Doe","Mary Johnson"),2) ; name
treatment <- rep(c("a","b"),each=3); treatment
result <- c("_","16","3","2","11","1"); result
tab3 <- data.frame(name,treatment,result)
```

## Table 1
```{r tab1, include=T}
knitr::kable(tab1, align = c("l","r","r"))
```

## Table 2
```{r tab2, include=T}
knitr::kable(tab2, align = c("l","r","r","r"))
```



## Tidy data: A standardized way to link structure of a dataset with its semantics  
- Data semantics
  - every value belongs to a *variable* and an *observation*
  - A variable contains all values that measure the same underlying attribute across units
  - An observation contains all values measured on the same unit across attributes

## Table 3
```{r tab3, include=T}
knitr::kable(tab3, align = c("l","l","r"))
```  
    
## Tidy data: A standardized way to link structure of a dataset with its semantics  
- Data semantics
  - it is easier to describe functional relationship between variables
  - it is easier to make comparison between group of comparison

|id|sex|weight|height|
|:---|:---|---:|---:|
|1|Male|65|168|
|2|Female|55|170|
|3|Female|38|168|
|4|Male|44|172|

## Tidy data
1.Each variable forms a column.
2.Each observation forms a row.
3.Each type of observational unit forms a table.  

```{r tabtidy, include=T}
knitr::kable(tab3, align = c("l","l","r"))
```

## Common problems of messy datasets
- Column headers are values not variable names.
- Multiple variables are stored in one column.
- Variables are stored in both rows and columns.
- Multiple types of observational units are stored in the same table.
- A single observational unit is stored in multiple tables.

**Use melting,string splitting, and casting**

## Column headers are values not variable names

|religion|<10k|10-20k|20-30k|30-40k|40-50k|50-75k|
|:---|---:|---:|---:|---:|---:|---:|
Agnostic|27|34|60|81|76|137|
Atheist|12|27|37|52|35|70|
Buddhist|27|21|30|34|33|58|
Catholist|418|617|732|670|638|1116|


## Melt the columns into rows

|religion|income|frequency|
|:---|:---|---:|
|Agnostic|<10k|27|
|Agonistic|10-20k|34|
|Agonistic|20-30k|60| 

## Multiple variables are stored in one column
|country|year|m014|m1524|m2534|f014|
|:---|---:|---:|---:|---:|---:|
|AD|2000|0|0|1|_|
|AE|2000|2|4|4|3|
|AF|2000|52|228|183|93|

## First, melt the data
|country|year|column|cases|
|:---|---:|:---:|---:|
|AD|2000|m014|0|
|AD|2000|m1524|10|
|AD|2000|m2534|12|
|AE|2000|f014|3|

## Then, split the string column
|country|year|sex|age|cases|
|:---|---:|:---|:---|---:|
|AD|2000|m|0-14|0|
|AD|2000|m|15-24|10|
|AD|2000|m|25-34|12|
|AE|2000|f|0-14|3|

## Variables are stored in both rows and columns
|id|year|month|element|d1|d2|d3|d4|d5|
|:---|:---:|---:|:---|---:|---:|---:|---:|---:|
|MX17004|2010|1|tmax|-|-|-|-|-|
|MX17004|2010|1|tmin|-|-|-|-|-|
|MX17004|2010|2|tmax|-|27.3|24.1|-|-|
|MX17004|2010|2|tmin|-|14.4|14.4|-|-|
|MX17004|2010|3|tmax|-|-|-|-|32.1|
|MX17004|2010|3|tmin|-|-|-|-|14.2|

## First melt the data  
|id|year|month|element|day|value|
|:---|:---:|:---:|---:|:---|---:|
|MX17004|2010|01|tmax|d30|27.8|
|MX17004|2010|01|tmin|d30|14.5|
|MX17004|2010|02|tmax|d2|27.3|
|MX17004|2010|02|tmin|d2|14.4|
|MX17004|2010|02|tmax|d3|24.1|
|MX17004|2010|02|tmin|d3|14.4|


## create date using year, month and day, and drop missing values
|id|date|element|value|
|:---|:---:|:---|---:|
|MX17004|2010-01-30|tmax|27.8|
|MX17004|2010-01-30|tmin|14.5|
|MX17004|2010-02-02|tmax|27.3|
|MX17004|2010-02-02|tmin|14.4|
|MX17004|2010-02-03|tmax|24.1|
|MX17004|2010-02-03|tmin|14.4|

## Next cast tmax and tmin

|id|date|tmax|tmin|
|:---|:---:|---:|---:|
|MX17004|2010-01-30|27.8|14.5|
|MX17004|2010-01-30|27.3|14.4|
|MX17004|2010-02-02|24.1|14.4|

## Multiple types of observational units are stored in the same table.
- e.g., 
  - volunteers/doctors and patients
  - clinic/hospital ward and patients
  - dental patients and teeth
  
## Song,artist and rank in one table
|year|artist|time|track|date|week|rank|
|:---:|:---|:---:|:---|:---:|---:|:---:|
|2000|2 Pac|4:22|Baby Don't Cry|2000-02-26|1|87|
|2000|2 Pac|4:22|Baby Don't Cry|200-03-04|2|82|
|2000|2Ge+her|3:15|The Hardest Part of...|2000-09-02|1|91|
|2000|2Ge+her|3:15|The Hardest Part of...|2000-09-09|2|87|
|2000|3 Doors Down|3:53|Kryptonite|2000-04-08|1|81|
|2000|3 Doors Down|3:53|Kryptonite|2000-04-15|2|70|

## Song, artist and rank in one table
- duplication with artists and times repeated
- break down into 
  - artist, song name and time
  - rank of song and week
- Relational data 
- Also require merging the two data set in one table using unique id 
  
## Song, artist and time
|id|artist|track|time|
|:---|:---|:---|:---:|
|1|2 Pac|Baby Don't Cry|4:22|
|2|2Ge+her|The Hardest part of ...|3:15|
|3|3 Doors Down|Kryptonite|3:53|
|4|3 Doors Down|Loser|4:24|

## Dates and rank of song
|id|date|rank|
|:---|:---:|---:|
|1|2000-02-26|87|
|1|2000-03-04|82|
|1|2000-03-11|72|
|1|2000-03-18|77|
|1|2000-03-25|87|
|1|2000-04-01|94|
|1|2000-04-08|99|
|2|2000-09-02|91|
|2|2000-09-09|87|
|2|2000-09-16|92|
|3|2000-04-08|81|
|3|2000-04-15|70|
|3|2000-04-22|68|
|3|2000-04-29|67|
|3|2000-05-06|66|

## One type in multiple table
- e.g., 
  - participants of survey 1 and 2
  - data of three quarterly reports
- Add a new column that records the original table
- Then append the tables


## Survey 1 table

|id|age|sex|Knowledge score|
|:---|---:|:---|---:|
|FHW01|34|M|70|
|FHW02|30|F|76|
|FHW03|38|F|69|

## Survey 2 table
|id|age|sex|Knowledge score|
|:---|---:|:---|---:|
|MPS01|39|M|78|
|MPS02|40|M|74|
|MPS03|38|F|70|

## Appended

|id|age|sex|Knowledge score|survey|
|:---|---:|:---|---:|---:|
|FHW01|34|M|70|1|
|FHW02|30|F|76|1|
|FHW03|38|F|69|1|
|MPS01|39|M|78|2|
|MPS02|40|M|74|2|
|MPS03|38|F|70|2|