---
title: "Sampling distribution"
author: "Kaung Myat Khant"
date: "2026-07-16"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = "")
```

## Definition of sampling distribution

-   Theoretical distribution\
-   A series of statistic from independent random samples of *same size* from *same population*\
-   Shape and spread depends on sample size (n) and variation in parent population

```{r getData, include=FALSE}
library(tidyverse)

# Set a seed so that our "random" population is exactly the same every time we run this code
set.seed(42)

# 1. Define the Gamma distribution parameters
# Target mean = 130. 
# We chose a shape of 20 to give us a good right-skew and realistic variance.
# Since Mean = shape * scale, then scale = Mean / shape.
target_mean <- 130
shape_param <- 20
scale_param <- target_mean / shape_param

# 2. Generate the population data frame
# We use tibble() to create a clean, modern data frame
population_data <- tibble(
    patient_id = 1:1000,
    sbp = rgamma(n = 1000, shape = shape_param, scale = scale_param)
)
population_data <- population_data |> 
    filter(sbp >= 90 & sbp <= 190) 
    

# 3. Calculate and save the exact population mean
pop_mean <- mean(population_data$sbp)
```

## Hypothetical population data

-   1000 population of \>50 years old women

-   Gamma distribution to get a right skewed distribution

-   Get the range of systolic blood pressure between 90mmHg and 190mmHg

-   The population mean is `r round(pop_mean)`.

## Distribution of the data

```{r distribution_population, include=TRUE}
# 4. Create a plot to visualize the right skew and the population mean line
ggplot(data = population_data, aes(x = sbp)) +
    geom_histogram(fill = "steelblue", color = "white", bins = 40) +
    # Add the vertical line for our population mean
    geom_vline(xintercept = pop_mean, color = "red", linetype = "dashed", linewidth = 1.2) +
    annotate(
        geom = "text", 
        x = pop_mean + 3,    # Places the text slightly to the right of the line
        y = 80,              # Sets the height of the text (adjust if your y-axis changes)
        label = paste("Mean =", round(pop_mean, 2)), 
        color = "red", 
        fontface = "bold",
        hjust = 0            # Left-aligns the text so it flows away from the line
    ) +
    coord_cartesian(xlim = c(90, 190)) +
    labs(
        title = "Hypothetical Population: Systolic Blood Pressure",
        subtitle = paste("908 Women over 50 | Population Mean =", round(pop_mean, 2), "mmHg"),
        x = "Systolic Blood Pressure (mmHg)",
        y = "Number of Women"
    ) +
    theme_minimal()
```

## Sampling

-   From this population of 908 women, **1000 samples** will be taken 3 ways
    1.  **5 women** in each sample
    2.  **30 women** in each sample
    3.  **100 women** in each sample
-   Then, we calculate the **sample statistic(mean)** of each sample and distribute along the X axis

## Sample Statistics

```{r sample, include=F}
# Define the number of simulations we want to run
n_sims <- 1000

# 1. Simulate the sampling process
# replicate() repeats the sampling process 1000 times.
# sample() draws 'size' number of blood pressures from our population data.
means_n5   <- replicate(n_sims, mean(sample(population_data$sbp, size = 5, replace = TRUE)))
means_n30  <- replicate(n_sims, mean(sample(population_data$sbp, size = 30, replace = TRUE)))
means_n100 <- replicate(n_sims, mean(sample(population_data$sbp, size = 100, replace = TRUE)))

# 2. Build individual data frames for each group
# This follows your exact "Group" and "Value" structure
df_pop <- tibble(
    Group = "1. Population",
    Value = population_data$sbp
)

df_n5 <- tibble(
    Group = "2. Sample Means (n=5)",
    Value = means_n5
)

df_n30 <- tibble(
    Group = "3. Sample Means (n=30)",
    Value = means_n30
)

df_n100 <- tibble(
    Group = "4. Sample Means (n=100)",
    Value = means_n100
)

# 3. Stack everything into a single master data frame
master_data <- bind_rows(df_pop, df_n5, df_n30, df_n100)

# Optional but recommended: Turn 'Group' into a factor to lock in the sorting order.
# Because you cleverly added "1., 2., 3., 4." to the names, it will naturally sort well, 
# but making it a factor ensures ggplot2 respects this exact order.
master_data <- master_data %>%
    mutate(Group = factor(Group))


```

```{r sampling_data, include=T}
# Take a quick peek at the structure of our new master dataset

master_data |> 
    group_by(Group) |> 
    summarise(Size = n(),
              Mean = mean(Value),
              Median = median(Value),
              SD = sd(Value),
              Q1 = quantile(Value, 0.25),
              Q3 = quantile(Value, 0.75),
              Min = min(Value),
              Max = max(Value)
              ) |> 
    knitr::kable(align = c("l","r","r","r","r","r","r","r","r"),
                 digits = 1)

```

## Sampling Distribution

-   When the sample size is large enough\
-   Distribution of the sample means follows **Normal distribution**\
-   **Mean of the samples' means** is equal to **True Population mean** ($\bar{X}=\mu$)

$$ \bar{X} = \frac{\sum_{i=1}^{n}{X_i}}{n}$$

## Distribution of Sample means

```{r distribution of sample mean, include=T, fig.width=8}
# Create the ultimate Central Limit Theorem visualization
ggplot(data = master_data, aes(x = Value)) +
    # Draw the histograms (we use a density stat so the y-axis represents proportions, 
    # which makes comparing different curve shapes much cleaner, though counts work too!)
    geom_histogram(fill = "steelblue", color = "white", bins = 40) +
    
    # Property A: Unbiasedness (The red dashed line for Population Mean)
    geom_vline(xintercept = pop_mean, color = "red", linetype = "dashed", linewidth = 1) +
    
    # Property B & C: The 2x2 Grid with a locked x-axis and free y-axis
    facet_wrap(~ Group, ncol = 2, scales = "free_y") +
    
    # Lock the x-axis view from 90 to 190 as requested
    coord_cartesian(xlim = c(90, 190)) +
    
    # Add clear labels for the presentation
    labs(
        title = "Demonstrating the Central Limit Theorem",
        subtitle = paste("Red dashed line = True Population Mean (", round(pop_mean, 2), " mmHg)", sep = ""),
        x = "Systolic Blood Pressure (mmHg)",
        y = "Frequency"
    ) +
    
    # A clean theme perfect for PowerPoint slides
    theme_minimal(base_size = 14) +
    theme(
        strip.text = element_text(face = "bold", size = 12), # Makes the 1, 2, 3, 4 labels pop
        panel.spacing = unit(1, "lines") # Adds a little breathing room between the 4 plots
    )
```


## Distribution of the estimates

-   *Mean of Sample means*$(\bar{X})$= `r round(pop_mean)`
-   Standard error of the mean (***SEM***) measures the variability of the distribution
-   If population SD is known, $SEM = \frac{\sigma }{\sqrt{n}}$
-   if only one sample SD is known, $SEM = \frac{s}{\sqrt{n}}$
-   Large *SEM* - Imprecise estimate
-   Small *SEM* - Precise estimate


## Standard error of the estimate
```{r SEM, include=TRUE}
sem <- data.frame(
    SEM=c(
        sd(population_data$sbp)/sqrt(length(population_data$sbp)),
        sd(sample(population_data$sbp, size = 5, replace = TRUE))/sqrt(5),
        sd(sample(population_data$sbp, size = 30, replace = TRUE))/sqrt(30),
        sd(sample(population_data$sbp, size = 100, replace = TRUE))/sqrt(100),
        sd(sample(population_data$sbp, size = 500, replace = TRUE))/sqrt(500)
    ),
    Mean=c(
        mean(population_data$sbp),
        mean(sample(population_data$sbp, size = 5, replace = TRUE)),
        mean(sample(population_data$sbp, size = 30, replace = TRUE)),
        mean(sample(population_data$sbp, size = 100, replace = TRUE)),
        mean(sample(population_data$sbp, size = 500, replace = TRUE))
    ),
    N = c(length(population_data$sbp),5,30,100,500), 
    ID = c("Population",as.character(1:4))
)
sem |> 
    select(ID,N, Mean, SEM) |> 
    knitr::kable(align = c("l","r","r","r"),
                 digits = 1)

```

## Central limits theorem

-   As sample size increases, the distribution of sample means approaches a normal curve regardless of the original population's shape.
-   The sampling distribution stays centred at the true population mean and its variance is $\frac{population\ variance}{sample\ size}$ 
-   Standard error decreases as sample size grows
-   The Central Limit Theorem and sampling distribution are foundation for inferential statistics.


## Use in inferential statistics

-   **Confidence Interval (CI)**: 
$$Estimator \pm Reliability\ Coefficient \times Standard\ Error$$

-   **Test Statistic (for P-values)**:

$$\frac{Relevant\ Statistic - Hypothesized\ Parameter}{Standard\ error}$$

# Thank you
